---
title: "Analysis"
output: html_document
---

```{r}
# Packages. 
library(tidyverse)
library(nnet)
# data 
data_clean <- read_csv("./data/bio_a_data_clean.csv") # this will be the name of the cleaned data. 
```

# Base Model. 

```{r}
# Fit a neural net multinomial regression. 
data_clean <- data_clean[-1,]
data_model <-  data_clean %>% 
  na.omit %>% 
  mutate(contr_or_not = ifelse(resp_contr_n == "No protection at all",0,1)) 
## ==== Logistic regression

# Model. 
glm_contr <- glm(contr_or_not ~ I(living_situation == "Greek") +
                   I(living_situation == "Off-campus") + 
                  resp_covid_safe 
                  #friend_covid_safe 
                  # race + 
                  # gender + 
                  # hispanic + 
                  # rel_status
                 , 
                 family = binomial(link = "logit"),
    data = data_model)

summary(glm_contr)
```

The coefficients show that there are a lot of people 

```{r}
## Multinomial 

sex_net <- multinom(resp_contr_n ~ I(living_situation == "Greek") + 
                      # Manually specify argument. 
                      I(living_situation == "Off-campus"), 
                    # Data argument. 
                    data = data_model)
sex_net %>% summary()

# make a table 
options(scipen = 999)
# make a coefficient table with P values 
z <- summary(sex_net)$coefficients/summary(sex_net)$standard.errors
# calculate Ps 
p <- (1 - pnorm(abs(z))) * 2
# calulate model table with p and z
m_table_1 <- data_frame("Variables" = sex_net$coefnames,"Two methods" = summary(sex_net)$coefficients[1,], 
           "Z-value " = z[1,],
           "P-value " = p[1,],
           "None at all" = summary(sex_net)$coefficients[2,],
           "Z-Value" = z[2,],
           "P-value" = p[2,]) %>% knitr:: kable(caption = "Table 2: Multinomial Regression Results: Base", digits = 3)
# NaN seem to be introduced because therre is separation in the data for one of them predictors. 
# Additionally, the numbers of non-binary individuals are so low that it is probably unlikely to get a response that is valid. 
```

In this multinomial logit model, the reference category is "On-campus".
In comparison to living on-campus: 
* Being in the category greek corresponds to a `r summary(sex_net)$coefficients[1,2]` decrease in the log odds of using two methods of contraception compared to one method, but it is not statistically significant. 
* Being in the category off-campus corresponds to a `r summary(sex_net)$coefficients[1,3]` decrease in the log-odds of using two methods of protection in comparison to using one method, but this is not statistically significant either. 

In comparison to those living on campus: 
* Being in the category greek corresponds to a `r summary(sex_net)$coefficients[2,2]` decrease in the log-odds of using no protection at all in comparison to one method, but this is not significant. 
* Being in the ctegory Off-campus, corresponds to an `r summary(sex_net)$coefficients[2,3]` decrease in the log-odds of using no protection at all in comparison to 1 method of protection, but this is not significant. 

# Full Model. 

```{r}
# Create a visual of that model. 
pred_base_greek <- predict(sex_net, newdata = data_frame(living_situation = "Greek"), type = "probs") 
pred_base_off <- predict(sex_net, newdata = data_frame(living_situation = "Off-campus"), type = "probs") 

```

